---
title: "Generating a csv with long form regional anomaly data from 2020 - 2025"
author: "Jay Spiros"
date: "`r Sys.Date()`"
output: html_document
---

# Setup and load the datasets

```{r Setup,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
library(tidyverse)
library(here) # cleanly construct paths to datasets
```

Annual temperature anomaly will come from the **NOAA National Centers for Environmental Information (NCEI) Global Time Series, mean temperature anomaly** regional datasets. These datasets use a baseline mean from 1910 - 2000 to calculate the temperature anomalies.

The csv file mapping continents/regions to countries was generated using Claude.ai, and is used to construct a dataframe where each row is a year, country, anomaly value

# Load and prepare datasets to be merged into one dataset

First define a helper function to load and get the datasets ready for merging.

```{r Define loading helper function}
load_and_trim_dataset <- function(region_to_load, 
                                  ds_date,
                                  min_year,
                                  max_year) {
    # returns a dataframe with the supplied region's anomaly data
    # trimmed to the years between min_year and max_year. 
    # Returned dataframe has columns "Year","Region"
        
    # load up the df from the dataset .csvs
    region_df <- read_csv(
        here("datasets/raw_datasets/temperature/global_anomaly",
            ds_date,
            paste0(
                "ncei-",
                tolower(region_to_load),
                "_temperature_anomaly-",
                ds_date,
                ".csv"
            )
        ),
        comment="#"
    )
    
    # filter the data to just include the years between min_year and max_year 
    # and change the "Anomaly" column to be the region's name so the dataframes
    # can be merged later
    # !! unquotes the string value in region_to_load, and := (Walrus operator)
    # is necessary to allow an expression to be used as the left hand side in 
    # functions from dplyr  
    trimmed_region_df <- 
        region_df %>% 
        filter(Year >= min_year & Year <= max_year) %>% 
        rename(!!region_to_load := Anomaly)
    
    return(trimmed_region_df)
}
```

```{r Load datasets,echo=TRUE,message=FALSE,warning=FALSE}

# datestamp on the datasets. must be a string.
dataset_date <- "202602"

# years to include in the output. must be be numeric
start_year <- 2020
end_year <- 2025

# global regions to work with
# North and South America need special treatment because of the space in their name
regions <- c("Africa", "Asia", "Europe", "namerica", "samerica", "Oceania", "Arctic", "Antarctic")

# dfs will be stored here to be merged later
dfs_to_merge <- list()

for (region in regions) {
    dfs_to_merge <- c(
        dfs_to_merge,
        list(load_and_trim_dataset(
                region_to_load=region, 
                ds_date=dataset_date, 
                min_year=start_year,
                max_year=end_year
            )
        )
    )
}

```


# Merge and check the new dataframe

```{r merge dataframes}
regional_anomaly <-
    reduce(dfs_to_merge, left_join, by="Year")
```

Quick check to make sure there are no missing values, and that the data types of each column are numeric.


```{r Check merged dataset}
colSums(is.na(regional_anomaly))
sapply(regional_anomaly, typeof)
```

# Pivot from wide to long data format

Long format will have `Year`, `Region` and `Anomaly` columns.

Also replace namerica with North America and samerica with South America once we have a `Regions` column.

```{r pivot, find and replace}
# the "!" is a logical NOT, so the pivot columns selection to pivot is 
# "all columns except Year"
# new long format df has columns "Year", "Region". "Anomaly" 
regional_anomaly_long <-
    regional_anomaly %>% 
    pivot_longer(!Year, names_to="Region", values_to="Anomaly")

# find and replace the column name friendly versions of North and South America
# with their full names
regional_anomaly_long <-
    regional_anomaly_long %>% 
        mutate(Region = 
                    recode(Region,
                        "namerica" = "North America",
                        "samerica" = "South America"
                    )
        )

```

# So tableau doesn't work with continents. 

The long format df created is still going to be useful for adding in all the countries of the world by ISO 3166 alpha-3 three letter country codes.

Had Claude AI generate a list of all countries, their ISO-3166 alpha-3 country code, and the continent they belong to.  
I moved Greenland (GRL), Svalbard (SJM), and Iceland (ISL) to the "arctic" so that region will be represented in the Tableau heat map.  
The continent of Antarctica is named "Antarctic" in the mapping file to match its name in the region file.

```{r Load mapping file,echo=TRUE,message=FALSE,warning=FALSE}
country_names <-
    read_csv(here("datasets/other_datasets/continent_to_country_mapping.csv"))
```

This code chunk builds the new dataframe by mapping the regional anomalies from the long format region file to the individual countries in those regions.

```{r Map regions to individual countries} 
country_anomaly_long <- regional_anomaly_long %>% 
    pmap(function(Year, Region, Anomaly) {
        country_list <- country_names %>% 
            filter(Continent == Region) %>% 
            pull(Country) %>% 
            as.list()
        pmap(list(country_list), function(country) {
            tibble(
                Year = Year,
                Country = country,
                Anomaly = Anomaly
            ) 
        }) %>% list_rbind()
    }) %>% list_rbind()
```

Another quick check

```{r Check expanded dataset}
colSums(is.na(country_anomaly_long))
sapply(country_anomaly_long, typeof)
```

# Output complete dataset to a .csv

```{r Write to CSV}

write_csv(country_anomaly_long, 
        here("datasets/processed_datasets",
            paste0(
                "ncei-by_country_anomaly_",
                 start_year,"_to_",end_year,"-",
                 dataset_date,
                 ".csv"
            )
        )
)
```


# Output new merged and pivoted dataset to a .csv
## Note this is not used, keeping for hostorical purposes.

```{r echo=FALSE,results='hide',message=FALSE,warning=FALSE}
write_csv(regional_anomaly_long, 
        here("datasets/processed_datasets",
            paste0(
                "ncei-merged_regional_anomaly_",
                 start_year,"_to_",end_year,"-",
                 dataset_date,
                 ".csv"
            )
        )
)
```

# These were used for testing and aren't needed, but could be useful in the future

```{r echo=FALSE,results='hide',message=FALSE,warning=FALSE}
new_df <- 
    tribble(
        ~Year, ~ Country, ~Continent, ~Anomaly,
        2025, "USA", "North America", 1.98,
        2025, "CAN", "North America", 1.98,
        2025, "MEX", "North America", 1.98,
        2025, "AUS", "Oceania", 1.16,
        2025, "NZL", "Oceania", 1.16,
        2025, "ATA", "Antarctica", 0.51,
        2025, "GRL", "Arctic", 2.72,
        2025, "SJM", "Arctic", 2.72,
        2025, "ISL", "Arctic", 2.72
    )
```


```{recho=FALSE,results='hide',message=FALSE,warning=FALSE}
new_df[1,]$Anomaly
```
