---
title: "Charting methane(CH<sub>4</sub>) atmospheric concentration increase over time"
author: "Jay Spiros"
date: "`r Sys.Date()`"
output: html_document
---

Methane(CH<sub>4</sub>) is a greenhouse gas of serious concern, as it is thought to trap even more heat than carbon dioxide(CO<sub>2</sub>). As the planet has warmed, more and more permafrost has started to melt, allowing methane trapped there to escape into the atmosphere and contribute to global warming.

# Setup and load the datasets

```{r Setup,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
library(tidyverse)
library(grid) # used to adjust margin in ggplot
library(here) # cleanly construct paths to datasets
```

Atmospheric CH<sub>4</sub> data will come from the **NOAA Global Monitoring Laboratory (GML) trends in atmospheric CH<sub>4</sub>, globally averaged from marine surface collection sites** annual mean dataset. CH<sub>4</sub> is measured in parts per billion (ppb).

```{r Load datasets,echo=TRUE,message=FALSE,warning=FALSE}

# not 100% necessary, but testing for some other datasets where it will be needed.
dataset_date <- "202603"
greenhouse_gas <- "CH4"

gml_annual_ch4_gl <- read_csv(
    here(
        "raw_datasets/greenhouse_gasses", 
        greenhouse_gas, 
        dataset_date, 
        paste0("gml-", tolower(greenhouse_gas), "_annmean_gl-", dataset_date, ".csv")
    ),
    comment = "#"
)

```

# Check the dataset

We're just doing a simple timeseries graph for CH<sub>4</sub>, so no need to do anything other than make sure the dataset has no missing or wrongly typed values.

The date range available is 1984 to 2025

Quick check to make sure there are no missing values, and that the data types of each column are numeric.

```{r Check merged dataset}
colSums(is.na(gml_annual_ch4_gl))
sapply(gml_annual_ch4_gl, typeof)
```

# Create visualizations

Plot the atmospheric CH<sub>4</sub> concentration 

```{r plot CH4 timeseries}
ch4_timeseriesplot <- ggplot(gml_annual_ch4_gl, mapping=aes(x=year, y=mean)) +
    geom_point(color="green") +
    geom_line(color="green") +
    labs(
        title="Atmospheric CH"[4]~"Concentration Over Time, 1984 to 2025",
        subtitle="CH"[4]~"concentration globally averaged from marine surface collection sites",
        x="Year",
        y="Atmospheric CH"[4]~"(ppb)",
        caption="Source: NOAA GML, trends in atmospheric CH4"
    ) +
    theme(
        plot.margin=unit(c(3, 4, 2, 4), "mm"),
        plot.caption=element_text(
            hjust=0,
            # vjust=-1
        )
    )
ch4_timeseriesplot
```

# Save plot to PNG file

```{r Save plots }
ggsave(here("images","ch4_timeseries.png"), plot=ch4_timeseriesplot, width = 9, height = 6)

```
