---
title: "Exploration of the correlation between annual CO<sub>2</sub> and global annual anomaly"
author: "Jay Spiros"
date: "`r Sys.Date()`"
output: html_document
---

# Setup and load the datasets

```{r Setup,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
library(tidyverse)
library(grid) # used to adjust margin in ggplot
library(here) # cleanly construct paths to datasets
library(patchwork) # stitch together plots onto one canvas
# library(ggh4x) # for changing the size of the plots inside the canvas (not currenlty used)
```

Atmospheric CO<sub>2</sub> data will come from the **NOAA Global Monitoring Laboratory(GML) trends in atmospheric CO<sub>2</sub>, Globally averaged from marine surface collection sites** annual mean dataset. CO<sub>2</sub> is measured in parts per million (ppm).

Annual temperature anomaly will come from the **NOAA National Centers for Environmental Information (NCEI) Global Time Series, mean temperature anomaly** global dataset. This dataset uses a baseline mean from 1901 - 2000 to calculate the temperature anomalies.

```{r Load datasets,echo=TRUE,message=FALSE,warning=FALSE}
gml_annual_co2_gl <- read_csv(
    here("datasets/raw_datasets/greenhouse_gasses/CO2/202603/gml-co2_annmean_gl-202603.csv"),
    comment = "#"
)
ncei_global_anomaly <- read_csv(
    here("datasets/raw_datasets/temperature/global_anomaly/202602/ncei-global_temperature_anomaly-202602.csv"),
    comment = "#"
)
```
x
# Organize and check the datasets

Merge the two datasets so we can chart the correlation between the annual global mean C0<sub>2</sub> levels and the annual anomaly.

The CO<sub>2</sub> levels dataset starts in 1979, so we will use data from 1979 - 2025.

The two datasets will be merged into a dataframe with the columns `Year`, `Anomaly` and `Mean_CO2`.

```{r Merge datasets}
global_anomaly_annual_co2_combined <- 
    ncei_global_anomaly %>% 
    filter(Year >= 1979) %>% 
    left_join(
        select(gml_annual_co2_gl, year, mean) %>% 
            rename(Year = year, Mean_CO2 = mean), 
        by = "Year"
    )
```

Quick check to make sure there are no missing values, and that the data types of each column are numeric.

```{r Check merged dataset}
colSums(is.na(global_anomaly_annual_co2_combined))
sapply(global_anomaly_annual_co2_combined, typeof)
```

# Create visualizations

Correlation of atmospheric CO<sub>2</sub> and global temperature anomaly

```{r plot correlation scatterplot,message=FALSE}
# string wrap annotation using stringr
wrapped_anomaly_ann <- 
    str_wrap(
        "Anomaly measured as deviation from 1901 to 2000 mean",
        width=30
    )


co2_anomaly_corrplot <- ggplot(global_anomaly_annual_co2_combined, mapping=aes(x=Mean_CO2, y=Anomaly)) +
   # geom_point(aes(color=Year)) +
    geom_point() +
    geom_smooth(se=FALSE) +
    labs(
        title="Atmospheric CO"[2]~"vs Global Mean Temperature Anomaly, 1979 to 2025",
        subtitle="C0"[2]~"concentration globally averaged from marine surface collection sites",
        x="Atmospheric CO"[2]~"(ppm)",
        y="Anomaly (\u00B0C)",
        caption="Sources:\nNOAA Global Monitoring Laboratory, trends in atmospheric CO2, Mar 2026\nNOAA National Centers for Environmental Information, Climate at a Glance: Global Time Series, Feb 2026"
    ) +
    annotate("text",
        x=350, y=1.00,
        label=wrapped_anomaly_ann,
        size=3.5,
        hjust=0.5 # centered
    ) +
    theme(
        plot.margin=unit(c(2, 3, 2, 3), "mm"),
        plot.caption=element_text(
            hjust=0,
            vjust=-1
        )
    )
    # To fiddle around with later: changes the dimensions of the plots
    # force_panelsizes(rows = unit(6, "cm"),
    #             cols = unit(10, "cm"))
co2_anomaly_corrplot
```

We'll create a CO<sub>2</sub> concentration and mean global temperature anomaly combo plot
First create the individual plots

```{r Plot CO2 timeseries}
co2_timeseriesplot <- ggplot(global_anomaly_annual_co2_combined, mapping=aes(x=Year, y=Mean_CO2)) +
    geom_point(color="deepskyblue") +
    geom_line(color="deepskyblue") +
    labs(
        title="Atmospheric CO"[2]~"Concentration Over Time, 1979 to 2025",
        subtitle="C0"[2]~"concentration globally averaged from marine surface collection sites",
        x="Year",
        y="Atmospheric CO"[2]~"(ppm)",
        # caption="Source: NOAA GML, trends in atmospheric CO2"
    ) +
    scale_y_continuous(position="right") +
    theme(
        plot.margin=unit(c(2, 3, 2, 3), "mm"),
        # plot.caption=element_text(
        #    hjust=0,
        # )
    )
    # To fiddle around with later: changes the dimensions of the plots
    # force_panelsizes(rows = unit(4, "cm"),
    #               cols = unit(8, "cm"))
co2_timeseriesplot
```
```{r Plot anomaly timeseries}
anomaly_timeseriesplot <- ggplot(global_anomaly_annual_co2_combined, mapping=aes(x=Year, y=Anomaly)) +
    geom_point(color="firebrick") +
    geom_line(color="firebrick") +
    labs(
        title="Global Mean Temperature Anomaly Over Time, 1979 to 2025",
        subtitle="Anomaly measured as deviation from 1901 to 2000 mean",
        x="Year",
        y="Anomaly (\u00B0C)",
        caption="Sources:\nNOAA Global Monitoring Laboratory, trends in atmospheric CO2, Mar 2026\nNOAA National Centers for Environmental Information, Climate at a Glance: Global Time Series, Feb 2026"
    ) +
    theme(
        plot.margin=unit(c(2, 3, 2, 3), "mm"),
        plot.caption=element_text(
            hjust=0,
            vjust=-1
        )
    )
    # To fiddle around with later: changes the dimensions of the plots
    # force_panelsizes(rows = unit(4, "cm"),
    #               cols = unit(8, "cm"))
anomaly_timeseriesplot
```

Then combine the two plots using the patchwork library

```{r Combined plots, message=FALSE, fig.width=14, fig.height=6}
combo_plot <- anomaly_timeseriesplot + co2_timeseriesplot
combo_plot
```

# Save plots to PNG files

```{r Save plots }
# correlation plot
ggsave(here("images","co2_v_globalanomaly.png"), plot=co2_anomaly_corrplot, width = 9, height = 6)

# combo plot
ggsave(here("images","co2_anomaly_combo.png"), plot=combo_plot, width = 14, height = 6)
```

