---
title: "Exploration of the correlation between annual CO<sub>2</sub> and global annual anomaly"
output:
  html_document:
    df_print: paged
---

# Setup and load the datasets

```{r Setup,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
library(tidyverse)
library(grid) # used to adjust margin in ggplot
library(here) # cleanly construct paths to datasets
```

Atmospheric CO<sub>2</sub> data will come from the **NOAA Global Monitoring Laboratory(GML) trends in atmospheric CO<sub>2</sub>, Globally averaged from marine surface collection sites** annual mean dataset. CO<sub>2</sub> is measured in parts per million (ppm).

Annual temperature anomaly will come from the **NOAA National Centers for Environmental Information (NCEI) Global Time Series, mean temperature anomaly** global dataset. This dataset uses a baseline mean from 1901 - 2000 to calculate the temperature anomalies.

```{r Load datasets,echo=TRUE,message=FALSE,warning=FALSE}
gml_annual_co2_gl <- read_csv(
    here("raw_datasets/greenhouse_gasses/CO2/202603/gml-co2_annmean_gl-202603.csv"),
    comment = "#"
)
ncei_global_anomaly <- read_csv(
    here("raw_datasets/temperature/global_anomaly/202602/ncei-global_temperature_anomaly-202602.csv"),
    comment = "#"
)
```

# Organize and check the datasets

Merge the two datasets so we can chart the correlation between the annual global mean C0<sub>2</sub> levels and the annual anomaly.

The CO<sub>2</sub> levels dataset starts in 1979, so we will use data from 1979 - 2025.

The two datasets will be merged into a dataframe with the columns `Year`, `Anomaly` and `Mean_CO2`.

```{r Merge datasets}
global_anomaly_annual_co2_combined <- 
    ncei_global_anomaly %>% 
    filter(Year >= 1979) %>% 
    left_join(
        select(gml_annual_co2_gl, year, mean) %>% 
            rename(Year = year, Mean_CO2 = mean), 
        by = "Year"
    )
```

Quick check to make sure there are no missing values, and that the data types of each column are numeric.

```{r Check merged dataset}
colSums(is.na(global_anomaly_annual_co2_combined))
sapply(global_anomaly_annual_co2_combined, typeof)
```

# Plot correlation scatterplot

```{r Generate plot,message=FALSE}
ggplot(global_anomaly_annual_co2_combined, mapping=aes(x=Mean_CO2, y=Anomaly)) +
    geom_point(aes(color=Year)) +
    geom_smooth(se=FALSE) +
    labs(
        title="Atmospheric CO"[2]~"vs Global Mean Temperature Anomaly, 1979 to 2025",
        subtitle="Anomaly measured as deviation from 1901 to 2000 mean",
        x="Atmospheric CO"[2]~"(ppm)",
        y="Anomaly (\u00B0C)",
        caption="Sources: NOAA GML, trends in atmospheric CO2\nNOAA NCEI Global Time Series mean temperature anomaly"
    ) +
    theme(
        plot.margin=unit(c(5, 5, 5, 5), "mm"),
        plot.caption=element_text(
            hjust=0,
            vjust=-1
        )
    )
```

