---
title: "Exploration of the correlation between atmospheric CO<sub>2</sub> and US average temperature"
author: "Jay Spiros"
date: "`r Sys.Date()`"
output: html_document
---

# Setup and load the datasets

```{r Setup,echo=TRUE,results='hide',message=FALSE,warning=FALSE}
library(tidyverse)
library(grid) # used to adjust margin in ggplot
library(here) # cleanly construct paths to datasets
```

Atmospheric CO<sub>2</sub> data will come from the **NOAA Global Monitoring Laboratory(GML) trends in atmospheric CO<sub>2</sub>, Mauna Loa Observatory** annual mean dataset. CO<sub>2</sub> is measured in parts per million (ppm).

US average annual temperature will come from the **NOAA National Centers for Environmental Information (NCEI), National Time Series average annual temperature 1895-2025** dataset. Average temp is measured in °F.

```{r Load datasets,echo=TRUE,message=FALSE,warning=FALSE}

# not 100% necessary, but testing for some other datasets where it will be needed.
dataset_date <- "202603"
greenhouse_gas <- "CO2"

gml_annual_co2_mlo <- read_csv(
    here(
        "datasets/raw_datasets/greenhouse_gasses", 
        greenhouse_gas, 
        dataset_date, 
        paste0("gml-", tolower(greenhouse_gas), "_annmean_mlo-", dataset_date, ".csv")
    ),
    comment = "#"
)

ncei_us_average_temp <- read_csv(
    here(
        "datasets/raw_datasets/temperature/us_national_average",
        dataset_date,
        paste0("ncei-continental_us_average_temperature-", dataset_date, ".csv")
    ),
    comment = "#"
)
```

# Organize and check the datasets

It's not strictly necessary to merge the datasets, but it feels cleaner.

The us temperature set has a `date` field that's **YYYYMM** so that will be separated into `year` and `month` fields. Column names will also be cleaned up.

The date range will be 1979 to 2025 to match the global mean anomaly correlation graph

```{r prep dataframes for merge}
ncei_us_average_temp_trimmed <-
    ncei_us_average_temp %>%  
    separate(col=Date, into=c("year", "month"), sep=4, convert=TRUE) %>% 
    filter(year >= 1979) %>% 
    rename(avg_temp = Value) %>% 
    select(year, avg_temp)

gml_annual_co2_mlo_trimmed <-
    gml_annual_co2_mlo %>% 
    rename(mean_co2 = mean) %>% 
    filter(year >= 1979) %>% 
    select(year, mean_co2)
```

```{r merge dataframes}
us_average_annual_co2_combined <-
    inner_join(ncei_us_average_temp_trimmed, gml_annual_co2_mlo_trimmed, by="year")
```

Quick check to make sure there are no missing values, and that the data types of each column are numeric.

```{r Check merged dataset}
colSums(is.na(global_anomaly_annual_co2_combined))
sapply(global_anomaly_annual_co2_combined, typeof)
```

# Create visualizations

Plot the atmospheric CO<sub>2</sub> vs the US national temperature average 

```{r plot correlation scatterplot,message=FALSE}
co2_ustemp_corrplot <- ggplot(us_average_annual_co2_combined, mapping=aes(x=mean_co2, y=avg_temp)) +
    geom_point(aes(color=year)) +
    scale_color_gradient(low = "green", high = "red") +
    geom_smooth(se=FALSE) +
    labs(
        title="Atmospheric CO"[2]~"vs US National Average Temperature, 1979 to 2025",
        subtitle="C0"[2]~"concentration collected at the Mauna Loa Observatory",
        x="Atmospheric CO"[2]~"(ppm)",
        y="US Average Temperature (\u00B0F)",
        caption="Sources:\nNOAA Global Monitoring Laboratory, trends in atmospheric CO2, Mar 2026\nNOAA National Centers for Environmental information, Climate at a Glance: National Time Series, Feb 2026"
    ) +
    theme(
        plot.margin=unit(c(3, 1, 3, 4), "mm"),
        plot.caption=element_text(
            hjust=0,
            vjust=-1
        )
    )
co2_ustemp_corrplot
```
For comparison, plot the US national average temperatures over time

```{r plot US average temp timeseries}
us_temp_timeseriesplot <- ggplot(us_average_annual_co2_combined, mapping=aes(x=year, y=avg_temp)) +
    geom_point(color="orange") +
    geom_line(color="orange") +
    labs(
        title="US National Average Temperature, 1979 to 2025",
        x="Year",
        y="US Average Temperature (\u00B0F)",
        caption="Source:\nNOAA National Centers for Environmental information, Climate at a Glance: National Time Series, Feb 2026"
    ) +
    theme(
        plot.margin=unit(c(3, 4, 3, 4), "mm"),
        plot.caption=element_text(
            hjust=0,
            # vjust=-1
        )
    )
us_temp_timeseriesplot
```

# Save plots to PNG files

```{r Save plots }
# correlation plot
ggsave(here("images","co2_v_us_average_temp.png"), plot=co2_ustemp_corrplot, width = 9, height = 6)

# combo plot
ggsave(here("images","us_average_temp.png"), plot=us_temp_timeseriesplot, width = 9, height = 6)
```
